\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-19 leslieandkenny.Rnw
Did some reorganizing, will work more out on paper and fill things in later

\noindent The model is
$$
y_i\sim\mathrm{overdispersed Binomial}(30, p_i, \omega),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\omega\) is an overdispersion parameter, so that the standard
deviation of the response in plot \(i\) is \(\sqrt{30\omega p_i(1-p_i)}\).
\end{itemize}

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \textbf{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. A separate analysis would
be needed for the more resistant population. \\

@@

\begin{figure}[h!]
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
#Warning message:
#In tableplot_checkCols(substitute(sortCol), colNames) :
#  Sorting on multiple columns not supported anymore. Only first column is taken.
tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status),
          nBins = 30, sortCol = "variety")
@
\end{figure}

\subsection{Model Fitting}
\label{model}

@@

<<glmms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
glmBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
              family = binomial, data = fert2)
# Warning message:
# glm.fit: fitted probabilities numerically 0 or 1 occurred
# Residual deviance: 427.939  on 125  degrees of freedom
# 427.939 / 125 = 3.423512

@@

glmmVarPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                       (1|block/variety) + (1|plot),
                     control = glmerControl(optimizer = "bobyqa",
                                            optCtrl = list("maxfun" = 20000)),
                     family = binomial, data = fert2)
# Warning messages:
# 1: In vcov.merMod(object, use.hessian = use.hessian) :
#   variance-covariance matrix computed from finite-difference Hessian is
# not positive definite or contains NA values: falling back to var-cov estimated from RX
# 2: In vcov.merMod(object, correlation = correlation, sigm = sig) :
#   variance-covariance matrix computed from finite-difference Hessian is
# not positive definite or contains NA values: falling back to var-cov estimated from RX
# plot          (Intercept) 1.03818  1.0189
# variety:block (Intercept) 0.28324  0.5322
# block         (Intercept) 0.00000  0.0000
# varietySNMS:n.trtLATE SPRING                -20.10372  8008.61084
# varietySNMS:status2CNTL                     -19.96154  7604.01229
# varietySNMS:n.trtEARLY SPRING:status2CNTL    20.93076  7604.01238
# varietySNMS:n.trtLATE SPRING:status2CNTL     20.61822 13745.14025
# deviance df.resid
#    726.8      147
# 726.8 / 147 = 4.9442177

@@

# Now with plot-level REs
cbind(coef(glmQBin)[c(1,7:13,34:55)], fixef(glmmVar))
cbind(arm::se.coef(glmQBin)[c(1,7:13,34:55)], arm::se.fixef(glmmVar))
# Pretty similar. Maybe this is ok?
@

@@

>>>>

\subsection{Potential Issues with Count Data}

\begin{figure}[h!]
\begin{center}
\caption{Illustration of separation.}
<<separation,echo=FALSE,out.width='0.6\\linewidth',fig.width=6>>=
curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
      xlab = expression(logit(p[i])), ylab = expression(p[i]))
segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)
@
\end{center}
\end{figure}

\noindent The following sections present examples of issues that could
arise while analyzing Nar's data. The examples all use simulated datasets.

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \emph{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. A separate analysis would
be needed for the more resistant population. \\

@@

\begin{figure}[h!]
\caption{Example \texttt{itableplot} output.}
<<tabplot,echo=FALSE,warning=FALSE,message=FALSE>>=
# Tableplot is confusing. Maybe omit this?
fert.sim1$v.trt <- with(fert.sim1, interaction(status2, n.trt, variety))
tableplot(fert.sim1, select = c(infected, variety, block, n.trt, status, v.trt),
          nBins = 30, sortCol = "v.trt")
@
\end{figure}

\noindent The model is
$$
y_i\sim\text{overdispersed Binomial}(30, p_i, \omega),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\omega\) is an overdispersion parameter, so that the variance
of the response in plot \(i\) is \(30\omega p_i(1-p_i)\).
\end{itemize}

\subsection{Binomial Generalized Linear Mixed Model}
\label{model2}

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
\begin{align*}
\mathrm{logit}(p_i)&=\mu+a_{b[i]}+(ab)_{b[i],v[i]}\\
&\quad+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}\\
&\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
+(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}\\
&\quad+(\beta\gamma\delta)_{v[i],t[i],j[i]}
\end{align*}
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(a_b\sim\mathrm{N}(0,\sigma^2_a)\) and
\((ab)_{b,v}\sim\mathrm{N}(0,\sigma^2_{ab})\) are random adjustments for
each block and each variety within each block.
\end{itemize}

\subsection{Fitting the GLM}
\label{model}

@@

<<glmms,echo=FALSE,warning=FALSE,message=FALSE,results='hide',cache=TRUE>>=
# Experimenting with different random effects.
# Are all block coefs 0 because they are nonidentified?

glmBin <- glm(cbind(infected, total) ~ block*variety + variety*n.trt*status2,
              family = binomial, data = fert2)
# Warning message:
# glm.fit: fitted probabilities numerically 0 or 1 occurred
# Residual deviance: 427.939  on 125  degrees of freedom
# 427.939 / 125 = 3.423512

@@

glmmVarPlot <- glmer(cbind(infected, total) ~ variety*n.trt*status2 +
                       (1|block/variety) + (1|plot),
                     control = glmerControl(optimizer = "bobyqa",
                                            optCtrl = list("maxfun" = 20000)),
                     family = binomial, data = fert2)
# Warning messages:
# 1: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
#   unable to evaluate scaled gradient
# 2: In checkConv(attr(opt, "derivs"), opt$par, ctrl = control$checkConv,  :
#   Model failed to converge: degenerate  Hessian with 2 negative eigenvalues# plot          (Intercept) 1.03818  1.0189
# variety:block (Intercept) 0.28324  0.5322
# block         (Intercept) 0.00000  0.0000
# varietySNMS:n.trtLATE SPRING                -20.10372  8008.61084
# varietySNMS:status2CNTL                     -19.96154  7604.01229
# varietySNMS:n.trtEARLY SPRING:status2CNTL    20.93076  7604.01238
# varietySNMS:n.trtLATE SPRING:status2CNTL     20.61822 13745.14025
# deviance df.resid
#    726.8      147
# 726.8 / 147 = 4.9442177

@@

# Now with plot-level REs
cbind(coef(glmQBin)[c(1,7:13,34:55)], fixef(glmmVar))
cbind(arm::se.coef(glmQBin)[c(1,7:13,34:55)], arm::se.fixef(glmmVar))
# Pretty similar. Maybe this is ok?
# Mark doesn't like it.
@

@@

\subsection{Fitting the GLMM}
